---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sweater

<!-- badges: start -->
<!-- badges: end -->

The goal of sweater (Simple Word Embedding Association TEst using R) is to test for biases in word embeddings.

## Installation

You can install the Github version of sweater with:

``` r
devtools::install_github("chainsawriot/sweater")
```

## Example

This example reproduces the detection of "Math. vs Arts" gender bias in Caliskan et al (2017).

```{r maths}
require(sweater)
data(glove_math) # a subset of the original GLoVE word vectors

S <- c("math", "algebra", "geometry", "calculus", "equations", "computation", "numbers", "addition")
T <- c("poetry", "art", "dance", "literature", "novel", "symphony", "drama", "sculpture")
A <- c("male", "man", "boy", "brother", "he", "him", "his", "son")
B <- c("female", "woman", "girl", "sister", "she", "her", "hers", "daughter")
sw <- sweater(glove_math, S, T, A, B)
sweater_es(sw)
```

The exact test described in Caliskan et al. (2017) is also available. But it takes a long time to calculate.

```r
## Don't do it. It takes a long time and is almost always significant.
exact_test(sw)
```

## References

1. Caliskan, Aylin, Joanna J. Bryson, and Arvind Narayanan. "Semantics derived automatically from language corpora contain human-like biases." Science 356.6334 (2017): 183-186.
