---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The original significance test proposed by Caliskan et al. in their paper is based on exact inference using a permutaion approach. This package provides such an exact test. But like any permutation test (such as Fisher's exact test), the number of permuations, as well as the running time, grows *factorially* --- O(n!) ---  with the number of samples (in the case, the total length of S and T). In the "math vs poetry" example used extensively in this package, the total length of S and T is 16. The total number of iterations needed is 16! = 20922789888000. The number grows 306x to 18! = 6402373705728000 by just adding one item each to S and T. It is unrealistic to expect an ordinary computer can do such computation within a reasonable time frame.

In the following benchmark, we studied how much time does it take to finish an exact test with the word set with a length of 2 to 4. I actually don't recommend trying anything beyond 6; it would take probably a few hours, days even. And actually, the function `weat_exact` will warn you about that.

```{r setup}
library(sweater)
S1 <- c("math", "algebra", "geometry", "calculus", "equations",
 "computation", "numbers", "addition")
T1 <- c("poetry", "art", "dance", "literature", "novel", "symphony", "drama", "sculpture")
A1 <- c("male", "man", "boy", "brother", "he", "him", "his", "son")
B1 <- c("female", "woman", "girl", "sister", "she", "her", "hers", "daughter")

do_exact_weat <- function(n = 4) {
    sw <- weat(glove_math, S1[1:n], T1[1:n], A1[1:n], B1[1:n])
    weat_exact(sw)
}

require(bench)
benchmark_res <- bench::mark(do_exact_weat(2),
                             do_exact_weat(3),
                             do_exact_weat(4),
                             relative = TRUE,
                             check = FALSE)
benchmark_res
```

Instead, I recommend using the resampling approximation of the exact inference. Actually, in the [Java code](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/DX4VWP/RQ1CIW&version=2.0) provided by Caliskan et al., the same resampling approximation is used.

```{r}
library(sweater)
set.seed(11111)
S1 <- c("math", "algebra", "geometry", "calculus", "equations", "computation", "numbers", "addition")
T1 <- c("poetry", "art", "dance", "literature", "novel", "symphony", "drama", "sculpture")
A1 <- c("male", "man", "boy", "brother", "he", "him", "his", "son")
B1 <- c("female", "woman", "girl", "sister", "she", "her", "hers", "daughter")

sw <- weat(glove_math, S1, T1, A1, B1)
weat_resampling(sw)
```


The method is extremely accurate. In the following simulation, we repeat the same resampling test for 50 times (which is still faster than conducting one exact test). The p-values are extremely close to the 0.018 repeated by Caliskan et al.

```{r}
res <- replicate(50, weat_resampling(sw)$p.value)
hist(res)
```

The precision can be increased by increasing the `n_resampling` parameter (default to 9999). In the Java code provided by Caliskan et al., this sets to 100000.

```{r}
res <- replicate(50, weat_resampling(sw, n_resampling = 15999)$p.value)
hist(res)
```

In the following example, we compare the exact inference with the resampling inference, when n = 5. On a regular computer, the exact test would take 1 minute.

```{r}
S2 <- c("math", "algebra", "geometry", "calculus", "equations")
T2 <- c("poetry", "art", "dance", "literature", "novel")
A2 <- c("male", "man", "boy", "brother", "he", "him", "his", "son")
B2 <- c("female", "woman", "girl", "sister", "she", "her", "hers", "daughter")

sw3 <- weat(glove_math, T2, S2, A2, B2)

system.time({exact_res <- weat_exact(sw3)})
```

```{r}
res <- replicate(100, weat_resampling(sw3)$p.value)
hist(res, main = "Distribution of p-values")
abline(v = exact_res$p.value, col = "red", lwd = 3)
```
